---
title: "BCB5460-R-Assignment"
author: "Emma Tran"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Inspection:

```{r, echo=FALSE}
genotypes <- read.table("fang_et_al_genotypes.txt", header = TRUE)
cat("Number of rows:", nrow(genotypes), "\nNumber of columns:", ncol(genotypes), "\nExamples of column names:", "\n")
head(genotypes)
```

- The columns are Sample_ID, JG_OTU, Group, and numerous SNP_IDs, with each SNP_ID representing a unique column.

```{r, echo=FALSE}
snp_pos <- read.table("snp_position.txt", header = TRUE, sep = "\t")
cat("Number of rows:", nrow(snp_pos), "\nNumber of columns:", ncol(snp_pos), "\nColumn names:", "\n")
snp_pos <- snp_pos[, c(1, 3, 4)]
head(snp_pos)
```

- The columns of snp_position.txt are SNP_ID, cdv_marker_id, Chromosome, Position, alt_pos, mult_positions, amplicon, cdv_map_feature.name, gene, candidate/random, Genaissance_daa_id, Sequenom_daa_id, count_amplicons, count_cmf, and count_gene.
- The Position column contains 'unknown' for missing values.
- The Chromosome column contains 'multiple' for entries with multiple values.


## Data Processing:
### Function to save files per chromosome:
```{r, echo=FALSE}
# Increasing position
save_chr_inc <- function(chr_id, merged, species) {
  if (!dir.exists(species)) dir.create(species)
  chr_data <- merged[merged$Chromosome == chr_id, ]
  chr_data$Position <- as.numeric(chr_data$Position)
  chr_data[is.na(chr_data)] <- "?"
  chr_data <- chr_data[order(chr_data$Position), ]
  write.table(chr_data, file.path(species, paste0(species, "_chr", chr_id, "_inc.txt")), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)}

# Decreasing position
save_chr_dec <- function(chr_id, merged, species) {
  if (!dir.exists(species)) dir.create(species)
  chr_data <- merged[merged$Chromosome == chr_id, ]
  chr_data$Position <- as.numeric(chr_data$Position)
  chr_data[is.na(chr_data)] <- "-"
  chr_data <- chr_data[order(chr_data$Position, decreasing = TRUE), ]
  write.table(chr_data, file.path(species, paste0(species, "_chr", chr_id, "_dec.txt")), sep = "\t", row.names = FALSE, col.names = TRUE, quote = FALSE)}
```

### Maize Data:
```{r, echo=FALSE}
library(gtools)
maize <- genotypes[genotypes[,3] %in% c("ZMMIL","ZMMLR","ZMMMR"), ]
transposed_maize <- as.data.frame(t(maize), stringsAsFactors = FALSE)
colnames(transposed_maize) <- NULL
maize_data <- transposed_maize
transposed_maize <- cbind(SNP_ID = rownames(maize_data), maize_data)
rownames(transposed_maize) <- NULL
maize_merged <- merge(snp_pos, transposed_maize, by = 1, sort = FALSE)
colnames(maize_merged)[4:ncol(maize_merged)] <- maize$Sample_ID
# write.table(maize_merged, "maize_genotypes_compilation.txt", sep="\t", row.names=FALSE, col.names=TRUE)
head(maize_merged)
```

```{r, echo=FALSE}
chromosomes <- unique(maize_merged$Chromosome)
species <- "maize"
invisible(lapply(chromosomes, save_chr_inc, merged = maize_merged, species = "maize"))
invisible(lapply(chromosomes, save_chr_dec, merged = maize_merged, species = "maize"))
```

### Teosinte Data:
```{r, echo=FALSE}
teosinte <- genotypes[genotypes[,3] %in% c("ZMPBA", "ZMPIL", "ZMPJA"), ]
transposed_teosinte <- as.data.frame(t(teosinte), stringsAsFactors = FALSE)
colnames(transposed_teosinte) <- NULL
teosinte_data <- transposed_teosinte
transposed_teosinte <- cbind(SNP_ID = rownames(teosinte_data), teosinte_data)
rownames(transposed_teosinte) <- NULL
teosinte_merged <- merge(snp_pos, transposed_teosinte, by = 1, sort = FALSE)
colnames(teosinte_merged)[4:ncol(teosinte_merged)] <- teosinte$Sample_ID
# write.table(maize_merged, "maize_genotypes_compilation.txt", sep="\t", row.names=FALSE, col.names=TRUE)
head(teosinte_merged)
```

```{r, echo=FALSE}
chromosomes <- unique(teosinte_merged$Chromosome)
species <- "teosinte"
invisible(lapply(chromosomes, save_chr_inc, merged = teosinte_merged, species = "teosinte"))
invisible(lapply(chromosomes, save_chr_dec, merged = teosinte_merged, species = "teosinte"))
```

## Visualization:
### SNPs per chromosome:
```{r, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)

maize_snp_counts <- maize_merged %>%
  group_by(Chromosome) %>%
  summarise(SNP_count = n())

ggplot(maize_snp_counts, aes(x = factor(Chromosome), y = SNP_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Maize", x = "Chromosome", y = "Number of SNPs")
```

```{r, echo=FALSE}
teosinte_snp_counts <- teosinte_merged %>%
  group_by(Chromosome) %>%
  summarise(SNP_count = n())

ggplot(teosinte_snp_counts, aes(x = factor(Chromosome), y = SNP_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Teosinte", x = "Chromosome", y = "Number of SNPs")
```


### Missing data and amount of heterozygosity
```{r, echo=FALSE}
genotype_cat <- function(gt) {
  if (is.na(gt) || gt == "?/?") {
    return("missing")
  } else if (grepl("^(A|C|G|T)/(A|C|G|T)$", gt)) {
    alleles <- strsplit(gt, "/")[[1]]
    if (alleles[1] == alleles[2]) {
      return("homozygous")
    } else {
      return("heterozygous")
    }
  } else {
    return("missing") 
  }
}
```

```{r, echo=FALSE}
sample_cols <- colnames(maize_merged)[4:ncol(maize_merged)]

maize_long <- maize_merged %>%
  select(all_of(sample_cols)) %>%
  pivot_longer(cols = everything(), names_to = "Sample", values_to = "Genotype") %>%
  mutate(Genotype_class = map_chr(Genotype, genotype_cat))

sample_summary <- maize_long %>%
  group_by(Sample, Genotype_class) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(prop = count / sum(count))

ggplot(sample_summary, aes(x = Sample, y = prop, fill = Genotype_class)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Maize", y = "Proportion", x = "Maize Samples", fill = "Genotype") +  
  theme_bw() +
  theme(axis.text.x = element_blank(),  
        axis.ticks.x = element_blank())
```


```{r, echo=FALSE}
sample_cols <- colnames(teosinte_merged)[4:ncol(teosinte_merged)]

teosinte_long <- teosinte_merged %>%
  select(all_of(sample_cols)) %>%
  pivot_longer(cols = everything(), names_to = "Sample", values_to = "Genotype") %>%
  mutate(Genotype_class = map_chr(Genotype, genotype_cat))

sample_summary <- teosinte_long %>%
  group_by(Sample, Genotype_class) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(prop = count / sum(count))

ggplot(sample_summary, aes(x = Sample, y = prop, fill = Genotype_class)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Teosinte", y = "Proportion", x = "Teosinte Samples", fill = "Genotype") +  
  theme_bw() +
  theme(axis.text.x = element_blank(),  
        axis.ticks.x = element_blank()) 
```

### My visualization:
```{r}
sample_cols <- colnames(maize_merged)[4:ncol(maize_merged)]

maize_long_chr <- maize_merged %>%
  select(Chromosome, all_of(sample_cols)) %>%
  pivot_longer(cols = all_of(sample_cols),
               names_to = "Sample",
               values_to = "Genotype") %>%
  mutate(Genotype_class = map_chr(Genotype, genotype_cat))

chromosome_summary <- maize_long_chr %>%
  group_by(Chromosome, Genotype_class) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Chromosome) %>%
  mutate(prop = count / sum(count))

ggplot(chromosome_summary, aes(x = Chromosome, y = prop, fill = Genotype_class)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Maize", y = "Proportion", x = "Chromosome", fill = "Genotype") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
sample_cols <- colnames(teosinte_merged)[4:ncol(teosinte_merged)]

teosinte_long_chr <- teosinte_merged %>%
  select(Chromosome, all_of(sample_cols)) %>%
  pivot_longer(cols = all_of(sample_cols),
               names_to = "Sample",
               values_to = "Genotype") %>%
  mutate(Genotype_class = map_chr(Genotype, genotype_cat))

chromosome_summary <- teosinte_long_chr %>%
  group_by(Chromosome, Genotype_class) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(Chromosome) %>%
  mutate(prop = count / sum(count))

ggplot(chromosome_summary, aes(x = Chromosome, y = prop, fill = Genotype_class)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Teosinte", y = "Proportion", x = "Chromosome", fill = "Genotype") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
















